---
title: "pmlAssignment.Rmd"
output: html_document
---

Goal

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively.

These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. 

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 
In this project, our goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. 

They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 
More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

Prepare the datasets
Read the training data

```{r}

require(caret)
require(data.table)
require(randomForest)
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
trainingData <- fread(url)
```

Read the testing data into a data table.

```{r}
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
testingData <- fread(url)
```

Variables without any missing values in the test dataset will be predictors.

```{r}
isMissing <- sapply(testingData, function (x) any(is.na(x) | x == ""))
isPredictor <- !isMissing & grepl("belt|[^(fore)]arm|dumbbell|forearm", names(isMissing))
predCandidates <- names(isMissing)[isPredictor]
predCandidates
```

Analyze the dataset to include only the predictor candidates and the variable - classe.

```{r}
varToInclude <- c("classe", predCandidates)
initialTrainingData <- trainingData[, varToInclude, with=FALSE]
dim(trainingData)
```

Split the dataset into a 60% training and 40% probing dataset.

```{r}
seed <- as.numeric(as.Date("2016-01-01"))
set.seed(seed)
inTrain <- createDataPartition(trainingData$classe, p=0.6)
finalTraining <- trainingData[inTrain[[1]]]
finalTesting <- trainingData[-inTrain[[1]]]
```

Preprocess the prediction variables by scaling.

```{r}
X <- finalTraining[, predCandidates, with=FALSE]
preProc <- preProcess(X)
preProc
prd <- predict(preProc, X)
cleanTrainingData <- data.table(data.frame(classe = finalTraining[, classe], prd))
```

Apply the centering and scaling to the probing dataset.

```{r}
X <- finalTesting[, predCandidates, with=FALSE]
prd <- predict(preProc, X)
cleanTestingData <- data.table(data.frame(classe = finalTesting[, classe], prd))
```

Check for near zero variance.
```{r}
nonZV <- nearZeroVar(cleanTrainingData, saveMetrics=TRUE)
if (any(nonZV$nonZV)) nonZV else message("zero variance variable not found.")
```

Train a prediction model.
Random forest enssures out of sample error is small. 
 
```{r} 
fittingData <- randomForest(classe ~. , data=cleanTrainingData)
prediction <- predict(fittingData, cleanTestingData, type = "class")
```
Model Evaluation
```{r} 
confusionMatrix(prediction, cleanTestingData$classe)
```
The model is 99% accurate on the training data as we have used a very efficient Random Forest ALgorithm.
